---
title: "Binary Classification Using Heart Failure Datasets"
description: "Example workflow predicting hospitalization using multimodal heart failure data."
---

This example shows a binary classification workflow for predicting hospitalization from multimodal heart failure data using `caretMultimodal`. The workflow covers loading the data, training modality-specific models and a stacked ensemble, and evaluating performance with ROC curves, contribution plots, and ablation analysis. The default settings on 5-fold cross validation are used to evaluate model performance.


# Load the data

The dataset used in this example is included in the `caretMultimodal` package. 

```{r, message = FALSE, warning = FALSE}
set.seed(123L)

# Load the data
data("heart_failure_datasets", package = "caretMultimodal")

# Inspect the available modalities
vapply(heart_failure_datasets, dim, integer(2))

# Grab the target column and drop the demo dataset
hospitalizations <- heart_failure_datasets$demo$hospitalizations
heart_failure_datasets$demo <- NULL
```

# Train models with caretMultimodal

```{r, message = FALSE, warning = FALSE}
# Set up hyperparameter tuning grid
alphas <- c(0.7, 0.775, 0.850, 0.925, 1)
lambdas <- seq(0.001, 0.1, by = 0.01)
tuneGrid <- expand.grid(alpha = alphas, lambda = lambdas)

# Train the base models
heart_models <- caretMultimodal::caret_list(
  target = hospitalizations,
  data_list = heart_failure_datasets,
  method = "glmnet",
  tuneGrid = tuneGrid
)

# Train the ensemble model
heart_stack <- caretMultimodal::caret_stack(
    caret_list = heart_models,
    method = "glmnet",
    tuneGrid = tuneGrid
)

```
# Evaluate and Interpret

## Model performance

```{r, message = FALSE, warning = FALSE}
summary(heart_stack)
```

**Note:** *Performance metrics returned by `summary.caret_stack()` are the resampling summaries generated by `caret` and may differ from metrics calculated directly from out-of-fold predictions. For example, the reported ROC value is the mean ROC across resampling folds, whereas `plot_roc.caret_stack()` calculates AUROC from the pooled out-of-fold predictions. See `compute_metric.caret_stack()` for calculating metrics directly from pooled out-of-fold predictions.*

```{r, message = FALSE, warning = FALSE}
# plot_roc is only for binary classification
caretMultimodal::plot_roc(heart_stack)

metric_fun <- function(preds, target) {
  pROC::roc(response = target, predictor = preds, quiet = TRUE)$auc
}

caretMultimodal::compute_metric(
  heart_stack, 
  metric_fun = metric_fun, 
  metric_name = "AUROC"
)
caretMultimodal::plot_metric(
  heart_stack, 
  metric_fun = metric_fun, 
  metric_name = "AUROC"
)
```

## Base model and feature contributions

```{r, message = FALSE, warning = FALSE}
caretMultimodal::compute_model_contributions(heart_stack)
caretMultimodal::plot_model_contributions(heart_stack)
caretMultimodal::compute_feature_contributions(heart_stack)
caretMultimodal::plot_feature_contributions(heart_stack)
```

## Ablation analysis

```{r, message = FALSE, warning = FALSE}
# Forward ablation
caretMultimodal::compute_ablation(
  heart_stack, 
  metric_fun = metric_fun, 
  metric_name = "AUROC"
)
caretMultimodal::plot_ablation(
  heart_stack, 
  metric_fun = metric_fun, 
  metric_name = "AUROC"
)

# Reverse ablation
caretMultimodal::compute_ablation(
  heart_stack, 
  metric_fun = metric_fun, 
  metric_name = "AUROC", 
  reverse = TRUE
)
caretMultimodal::plot_ablation(
  heart_stack, 
  metric_fun = metric_fun, 
  metric_name = "AUROC", 
  reverse = TRUE
)
```
