---
title: "Regression Using Pregnancy Datasets"
description: "Example workflow predicting gestational age using 7 related datasets."
---

This example shows a regression workflow for predicting gestational age from multimodal pregnancy data using caretMultimodal. The workflow covers downloading and preprocessing seven paired omics assays, training modality-specific elastic net models and a stacked ensemble under leave-one-subject-out cross-validation, and evaluating performance with Spearman correlation, contribution plots, and ablation analysis.

# Load and preprocess the data

```{r load-and-preprocess, cache = TRUE}
# Download the data
url <- "https://nalab.stanford.edu/wp-content/uploads/termpregnancymultiomics.zip"
buf <- curl::curl_fetch_memory(url)$content
load(archive::archive_read(rawConnection(buf), file = "termpregnancymultiomics/Data.Rda"))

# Name the input data
names(InputData) <- c("Cellfree RNA", "Plasma Luminex", "Serum Luminex", "Microbiome", "Immune System", "Metabolomics", "Plasma Somalogic")

# Correct nonstandard feature names
InputData <- lapply(InputData, function(df) {
  df <- as.data.frame(df)
  names(df) <- make.names(iconv(names(df), from = "latin1", to = "UTF-8", sub = ""), unique = TRUE)
  df
})

names(InputData$Metabolomics) <- paste("Metabolomic Feature", seq_along(names(InputData$Metabolomics)))

# Remove postpartum samples
postpartum_indices <- which(featureweeks < 0)
InputData <- lapply(InputData, function(x) x[-postpartum_indices, ])
featurepatients <- featurepatients[-postpartum_indices]
featureweeks <- featureweeks[-postpartum_indices]

```

# Train models with caretMultimodal

```{r train-models, cache = TRUE}
set.seed(42L)

# Define hyperparameter tuning grid
alphas <- seq(0, 1, 0.1)
lambdas <- seq(0, 12, by = 0.5)
tuneGrid <- expand.grid(alpha = alphas, lambda = lambdas)

# Set up leave-one-subject out cross validation
loso_folds <- lapply(unique(featurepatients), function(p) {
  which(featurepatients != p)  # indices of all samples NOT in this patient
})
 
trControl <- caret::trainControl(
  method = "cv",
  index = loso_folds,
  savePredictions = "final",
  summaryFunction = caret::defaultSummary
)
 
# Train the base models
pregnancy_models <- caretMultimodal::caret_list(
  data_list = InputData,
  target = featureweeks,
  method = "glmnet",
  tuneGrid = tuneGrid,
  trControl = trControl,
  trim = FALSE
)
 
# Train the ensemble model
pregnancy_stack <- caretMultimodal::caret_stack(
  pregnancy_models,
  method = "glmnet",
  tuneGrid = tuneGrid,
  trControl = trControl
)
```

# Evaluate and Interpret

```{r}
summary(pregnancy_stack)
caretMultimodal::plot_model_contributions(pregnancy_stack)
caretMultimodal::plot_feature_contributions(pregnancy_stack)

metric_fun <- function(preds, target) {
  -log10(cor.test(preds, target, method = "spearman")$p.value)
}

caretMultimodal::plot_metric(pregnancy_stack, metric_fun = metric_fun, metric_name = "-log(p-val)")
caretMultimodal::plot_ablation(pregnancy_stack, metric_fun = metric_fun, metric_name = "-log(p-val)")
```
