---
title: "Spatial Transcriptomics Heart Transplant Rejection Modeling"
description: "Construct multimodal pseudobulk views from Xenium spatial transcriptomics for modeling with caretMultimodal."
---

This vignette demonstrates a preprocessing workflow for deriving multimodal sample-level views from spatial single-cell transcriptomics data and using them with `caretMultimodal`. 

Starting from Xenium heart transplant biopsy data from GEO (GSE290577), we aggregate single-cell expression into pseudobulk profiles stratified by cell type and sample, normalize each view independently, and assemble the result into a MultiAssayExperiment suitable for multimodal modeling. The resulting object can be used directly with caretMultimodal to train modality-specific learners and stacked ensembles. For simplicity, the modeling example shown here focuses on a binary classification task comparing ellular versus antibody mediated rejection.

# Load and Preprocess the data

### Collect the data from GEO
```{r download-data, cache=TRUE}
# Load the xenium data (takes a minute or two)
xenium_url <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE290nnn/GSE290577/suppl/GSE290577_heart_spatial.rds"
xenium_raw <- curl::curl_fetch_memory(xenium_url)$content
xenium_obj <- unserialize(xenium_raw)
print(xenium_obj)

# Load the metadata
meta_url <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE290nnn/GSE290577/suppl/GSE290577_heart_spatial_metadata.txt.gz"
meta_raw <- curl::curl_fetch_memory(meta_url)$content
meta_df <- read.delim(
    text = rawToChar(memDecompress(meta_raw, type = "gzip")),
    sep = "\t",
    header = TRUE,
    stringsAsFactors = FALSE
  )
# Restrict to pre-biopsy, non-mixed rejection samples
meta_df <- meta_df |>
  dplyr::filter(biopsy_timing == "pre") |>
  dplyr::filter(biopsy_rejection_type != "mixed_rejection") |>
  dplyr::arrange(patient_id)

head(meta_df)
```

### Prepare cell-level metadata and count matrix

```{r prepare-inputs}
# Extract per-cell metadata and restrict to retained samples
cell_meta <- xenium_obj@meta.data |>
  dplyr::filter(Sample %in% meta_df$Sample)

# Standardize fields required by muscat
cell_meta <- cell_meta |>
  dplyr::mutate(
    response = biopsy_rejection_type,
    id = Sample,
    group = patient_id,
    sample_id = Sample,
    cluster_id = ct_second_pass
  )

# Extract raw counts (genes x cells), aligned to filtered metadata
rna_counts <- xenium_obj@assays$RNA@counts[, rownames(cell_meta), drop = FALSE]
```

### Construct SingleCellExperiment

```{r make-sce}
sce_rna <- SingleCellExperiment::SingleCellExperiment(
  assays = list(counts = rna_counts),
  colData = S4Vectors::DataFrame(cell_meta)
)

sce_rna
```

### Pseudobulk by cell type and sample with Muscat

```{r pseudobulk}
pb_rna <- muscat::aggregateData(
  sce_rna,
  assay = "counts",
  fun = "sum",
  by = c("cluster_id", "sample_id")
)

pb_rna
```

### Normalize pseudobulk assays (TMM + logCPM)

```{r normalize}
for (assay_name in SummarizedExperiment::assayNames(pb_rna)) {
  mat <- SummarizedExperiment::assay(pb_rna, assay_name)
  lib_size <- colSums(mat)

  keep_samples <- lib_size > 0

  norm_mat <- matrix(
    0,
    nrow = nrow(mat),
    ncol = ncol(mat),
    dimnames = dimnames(mat)
  )

  if (sum(keep_samples) >= 2) {
    dge <- edgeR::DGEList(counts = mat[, keep_samples, drop = FALSE])
    dge <- edgeR::calcNormFactors(dge, method = "TMM")

    norm_mat[, keep_samples] <- edgeR::cpm(
      dge,
      log = TRUE,
      prior.count = 0.5
    )
  }

  SummarizedExperiment::assay(pb_rna, paste0(assay_name, "_logCPM")) <- norm_mat
}

pb_rna_norm <- pb_rna
pb_rna_norm
```

### Convert to MultiAssayExperiment

```{r build-mae}
# Identify normalized assays
log_assays <- grep("_logCPM$", SummarizedExperiment::assayNames(pb_rna_norm), value = TRUE)

# Sample-level metadata
sample_meta <- as.data.frame(SummarizedExperiment::colData(pb_rna_norm))
rownames(sample_meta) <- colnames(pb_rna_norm)
sample_meta <- S4Vectors::DataFrame(sample_meta)

# Build one SummarizedExperiment per cell type
se_list <- lapply(log_assays, function(assay_name) {
  mat <- SummarizedExperiment::assay(pb_rna_norm, assay_name)

  keep_genes <- matrixStats::rowVars(mat, na.rm = TRUE)
  keep_genes <- is.finite(keep_genes) & keep_genes > 0
  mat <- mat[keep_genes, , drop = FALSE]

  SummarizedExperiment::SummarizedExperiment(
    assays = list(logCPM = mat),
    colData = sample_meta[colnames(mat), , drop = FALSE]
  )
})

names(se_list) <- sub("_logCPM$", "", log_assays)

# Map assay columns to primary sample IDs
sample_map <- do.call(rbind, lapply(names(se_list), function(exp_name) {
  sample_ids <- colnames(se_list[[exp_name]])

  data.frame(
    assay = exp_name,
    primary = sample_ids,
    colname = sample_ids,
    stringsAsFactors = FALSE
  )
}))

# Construct MAE
mae_pb <- MultiAssayExperiment::MultiAssayExperiment(
  experiments = MultiAssayExperiment::ExperimentList(se_list),
  colData = sample_meta,
  sampleMap = sample_map
)

mae_pb
```

### Summarize and filter low-coverage experiments

```{r experiment-filter}
# Extract sample-level response labels from the MAE
response_vec <- SummarizedExperiment::colData(mae_pb)$response
names(response_vec) <- rownames(SummarizedExperiment::colData(mae_pb))
response_levels <- sort(unique(as.character(response_vec)))

# Minimum number of nonzero samples required per response group
min_samples <- 15

# Count nonzero-library samples in each response group for every assay
experiment_summary <- lapply(names(MultiAssayExperiment::experiments(mae_pb)), function(exp_name) {
  se <- MultiAssayExperiment::experiments(mae_pb)[[exp_name]]
  lib_size <- colSums(SummarizedExperiment::assay(se, "logCPM"))

  nonzero_samples <- names(lib_size)[lib_size > 0]
  nonzero_tab <- table(factor(response_vec[nonzero_samples], levels = response_levels))

  data.frame(
    experiment = exp_name,
    response = response_levels,
    n_nonzero_lib = as.integer(nonzero_tab),
    stringsAsFactors = FALSE
  )
}) |>
  dplyr::bind_rows()

# Retain assays with at least min_samples nonzero samples in every response group
keep_names <- experiment_summary |>
  dplyr::group_by(experiment) |>
  dplyr::summarise(keep = all(n_nonzero_lib >= min_samples), .groups = "drop") |>
  dplyr::filter(keep) |>
  dplyr::pull(experiment)

# Subset the MAE to retained assays
mae_pb_keep <- mae_pb[, , keep_names]

mae_pb_keep
```

### Visualize per-assay logCPM distributions

```{r plot-distributions}
plot_df <- lapply(names(MultiAssayExperiment::experiments(mae_pb_keep)), function(exp_name) {
  mat <- SummarizedExperiment::assay(
    MultiAssayExperiment::experiments(mae_pb_keep)[[exp_name]],
    "logCPM"
  )

  data.frame(
    value = as.vector(mat),
    experiment = exp_name
  )
}) |>
  dplyr::bind_rows()

ggplot2::ggplot(plot_df, ggplot2::aes(x = value)) +
  ggplot2::geom_histogram(bins = 50) +
  ggplot2::facet_wrap(~ experiment, scales = "free") +
  ggplot2::theme_bw() +
  ggplot2::theme(
    strip.text = ggplot2::element_text(size = 8),
    axis.text = ggplot2::element_text(size = 6)
  )
```

# Train models with caretMultimodal

There are several ways to model this data. For simplicity, this example uses a binary classification task comparing cellular versus antibody mediated rejection (ignoring subjects with no rejection).

```{r train-models, cache=TRUE}
set.seed(42L)

data_list <- caretMultimodal::prepare_mae(mae_pb_keep, transpose = TRUE)

# caretMultimodal currently expects syntactically safe view names
view_labels <- names(data_list)
view_names <- make.names(view_labels, unique = TRUE)
names(data_list) <- view_names

# Restrict to AMR vs CMR samples
responses <- factor(trimws(SummarizedExperiment::colData(mae_pb_keep)$response))
keep <- responses %in% c("antibody_rejection", "cellular_rejection")

data_list <- lapply(data_list, function(x) x[keep, , drop = FALSE])
response_subset <- droplevels(responses[keep])

# Elastic net hyperparameter grid
tuneGrid <- expand.grid(
  alpha = seq(0, 1, by = 0.1),
  lambda = 10^seq(-4, 1, length.out = 50)
)

# Repeated cross-validation
trControl <- caret::trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 5,
  savePredictions = "final",
  classProbs = TRUE,
  summaryFunction = caret::twoClassSummary
)

# Train base models
htr_models <- caretMultimodal::caret_list(
  target = response_subset,
  data_list = data_list,
  method = "glmnet",
  tuneGrid = tuneGrid,
  trControl = trControl
)

# Train stacked ensemble
htr_stack <- caretMultimodal::caret_stack(
  caret_list = htr_models,
  method = "glmnet",
  tuneGrid = tuneGrid,
  trControl = trControl
)
```
# Evaluate Models

```{r evaluate-models}
summary(htr_stack)
caretMultimodal::plot_roc(htr_stack)
caretMultimodal::plot_model_contributions(htr_stack)
caretMultimodal::plot_feature_contributions(htr_stack)
```
